Disturbance-aware minimum-time planning strategies for motorsport vehicles with probabilistic safety certificates

### 2025-06-04 closed_loop_planning.tex
[mg] general reviews

The approach is based on the three following steps: (i) A nominal time-optimal trajectory is planned: this will be referred to as nominal feed-forward optimal trajectory. This represents the trajectory that an expert driver would follow in an ideal scenario without uncertainty. In this step, the mean trajectory coincides with the actual path that the vehicle would follow under deterministic dynamics.

(ii) A closed-loop controller mimicking the driver's action is computed which stabilizes the nominal feed-forward optimal trajectory computed at step (i). We assume a discrete time-varying Linear Quadratic Regulator (LQR) controller is a good approximation of how an expert driver can track a prescribed trajectory.

(iii) In the last step the stochastic framework is reintroduced and the time-optimal planning problem is robustified using the Lyapunov framework. The main idea is to \emph{re-plan} a mean trajectory of a stochastic system such that, a time-varying LQR controller that aims at stabilizing it, is able to properly tame the propagation of the uncertainty. This amounts to ensuring that the robustified path constraints are satisfied. More specifically, this considers two key aspects: firstly, that the covariance matrix is propagated accounting explicitly for the controlled system dynamics; secondly, that the path constraints are robustly satisfied by incorporating the covariance estimation propagated through the closed-loop controlled system. Some aspects that may appear a bit technical are in fact essential. In this third step, not only are states and controls re-planned, but also the time-varying LQR gains. As a matter of fact, they need to explicitly account for the changed requirements associated with stabilizing a trajectory different from the nominal one, which must now satisfy the robustified constraints. 
\subsection{Step 1 -- Nominal feed-forward time-optimal trajectory planning}
\label{sec:nominalFF}
Step (i) consists in planning a nominal feed-forward trajectory. This assumes deterministic system dynamics, where $\hbmu_k$, $\hbu_k$, and $\hbz_k$ denote differential states, controls, and algebraic states, respectively, for \( k = 0, \ldots, N \). To avoid redundancy, we note that the NLP to be solved is a special case of~\eqref{eq:DOCP}, where~\eqref{eq:DOCPdP},~\eqref{eq:DOCPdPIC}, and the backoff terms are omitted, since the goal is to compute a nominal trajectory.

@@

\subsection{Step 3 -- Closed-loop robustified planning}
\label{sec:CLrobustified}
%\begin{align}
%    \underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k, \bK_k}{\text{minimize}} \quad & J_k(\bmu_k,\bxi_k, \bu_k) \label{eq:ROCPcost}\\
%       \text{s.t.}\notag\\
%          & f_0(x) - t \le 0 \quad \phantom{i=1, \dots, m}\\
%          & f_i(x) \le 0, \quad i=1, \dots, m \quad i=1, \dots, m\\
%          & h_i(x) = 0,   \quad i=1, \dots, p,
%\end{align}
As the third and last step, a closed-loop robustified planning problem is set up. Its aim is the definition of a mean trajectory of a stochastic system such that, working in tandem with state-feedback controller that aims at stabilizing it, is able to mitigate the perturbations and reduce the propagation of uncertainty. This helps to guarantee that the robustified path constraints can be satisfied over the entire planning horizon. The discretized version of this problem takes the form of the following NLP:
\begin{subequations}\label{eq:ROCP}
\begin{alignat}{3}
\underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k, \bK_k}{\text{minimize}}
    & \quad J_k(\bmu_k,\bxi_k, \bu_k) & & \label{eq:ROCPcost} \\
\text{s.t.}\notag\\
\bzero      & = \; \bPsimu_k(\bmu_{k-1},\bmu_k,\bxi_k, \bu_k,\bz_k),
\quad k = 1,\ldots, N \label{eq:ROCPdyn} \\
\bmu_0      & = \; \bar{\bmu}_0 \label{eq:ROCPdynIC} \\
\bzero      & = \; \bPsiPCL_k(\bmu_k,\bxi_k, \bu_k, \tbP_{k-1},\tbP_k,\bSi_k,\bz_k, \bK_k),
\quad k = 1,\ldots, N \label{eq:ROCPdP} \\
\tbP_0       & = \; \bar{\bP}_0 \succeq 0 \label{eq:ROCPdPIC} \\
\bzero      & = \; \bOm_k(\bmu_k,\bxi_k, \bu_k,\bz_k), \quad k = 0,\ldots, N \label{eq:ROCPpath} \\
0    & \geq  h_i(\bmu_k, \bu_k, \bz_k) + \be_i(\bmu_k, \bu_k, \tbP_k, \bz_k), \quad k = 1,\ldots, N;\; \quad i \in \calI \label{eq:ROCPconstraints}\\
-\de \bK_k   & \leq \bK_k - \hbK_k \leq  \de \bK_k \label{eq:ROCPdeltaK}, \quad k = 0,\ldots, N
\end{alignat}
\end{subequations}
Here,~\eqref{eq:ROCPdP} denotes the discrete-time covariance dynamics under the stabilizing state-feedback controller \( \bbu_k = -\bK_k \bbmu_k \). To distinguish this case from~\eqref{eq:DOCP}, we use the symbol \( \tbP \) to denote the closed-loop (CL) covariance matrix, whose discrete evolution corresponds to a collocation-based approximation of the continuous-time dynamics $\dot{\tbP}(t) = \tbA(t) \tbP(t) + \tbP(t)\tbA(t)^T + \bQ(t)$, with $\tbA(t)=\bA(t)-\bB(t)\bK(t)$. It is worth noting that, accordingly, in equation~\eqref{eq:ROCPconstraints} the backoff term $\be_i$ is a function of closed-loop uncertainty $\tbP_k$. Another key aspect is that the state-feedback matrix $\bK_k$ is also treated as a decision variable. This reflects the fact that the driver may adjust his action to stabilize the new (to-be-planned) nominal trajectory using a policy that differs from the previous $\hbK_k$, computed in step (ii) via LQR. To promote convergence, however, $\bK_k$ is restrained within prescribed bounds ($\de \bK_k$) relative to $\hbK_k$. In our tests $\de \bK_k = 0.1 \hbK_k$.

@@

To correctly enforce the mutual exclusivity within the LQR synthesis via Riccati recursion, the state-feedback gain must be computed with respect to the minimal control input. Specifically, the control input matrix $\tbB$ is employed and the gain $\tilde{\bK} \in \mathbb{R}^{2 \times 6}$ is sought for which operates on the independent variation \(\delta\tilde{\bu}\). This is structured as
\[
\tilde{\bK} = \begin{bmatrix} \tilde{\bK}_F^T & \tilde{\bK}_\delta^T \end{bmatrix}^T, \quad \tilde{\bK}_F, \tilde{\bK}_\delta \in \mathbb{R}^{1 \times 6}.
\]
The extended gain $\bK$ is automatically recovered via $\bK = \bH \tbK$ and, depending on the operating mode, it will result in the following expressions:
\begin{itemize}
  \item during \emph{acceleration}: \(\bK = \begin{bmatrix} \tilde{\bK}_F^T & \mathbf{0}^T & \tilde{\bK}_\delta^T \end{bmatrix}^T\),
  \item during \emph{braking}: \(\bK = \begin{bmatrix} \mathbf{0}^T & \tilde{\bK}_F^T & \tilde{\bK}_\delta^T \end{bmatrix}^T\).
\end{itemize}
This formulation preserves consistency with the complementarity constraint while retaining a structure amenable to controller synthesis.

>>>>

The approach is based on the three following steps. 
\begin{enumerate}[label=\roman*)]
\item A nominal time-optimal trajectory is planned: this will be referred to as nominal feed-forward optimal trajectory. This represents the trajectory that an expert driver would follow in an ideal scenario without uncertainty. In this step, the mean trajectory coincides with the actual path that the vehicle would follow under deterministic dynamics.

\item A closed-loop controller mimicking the driver's action is computed to stabilize the nominal feed-forward optimal trajectory obtained at step (i). We assume a discrete time-varying Linear Quadratic Regulator (LQR) controller is a good approximation of how an expert driver can track a prescribed trajectory.

\item In the last step the stochastic framework is reintroduced and the time-optimal planning problem is robustified using the Lyapunov framework. The main idea is to \emph{re-plan} a mean trajectory of a stochastic system such that, a time-varying LQR controller that aims at stabilizing it, is able to properly tame the propagation of the uncertainty. This amounts to ensuring that the robustified path constraints are satisfied. More specifically, this considers two key aspects: firstly, that the covariance matrix is propagated accounting explicitly for the controlled system dynamics; secondly, that the path constraints are robustly satisfied by incorporating the covariance estimation propagated through the closed-loop controlled system. Some aspects that may appear a bit technical are in fact essential. In this third step, not only are states and controls re-planned, but also the time-varying LQR gains. As a matter of fact, they need to explicitly account for the changed requirements associated with stabilizing a trajectory different from the nominal one, which must now satisfy the robustified constraints. 
\end{enumerate}

\subsection{Step 1 -- Nominal feed-forward time-optimal trajectory planning}
\label{sec:nominalFF}
Step (i) consists of planning a nominal feed-forward trajectory. This assumes deterministic system dynamics, where $\hbmu_k$, $\hbu_k$, and $\hbz_k$ denote differential states, controls, and algebraic states, respectively, for \( k = 0, \ldots, N \). To avoid redundancy, we note that the NLP to be solved is a special case of~\eqref{eq:DOCP}, where~\eqref{eq:DOCPdP},~\eqref{eq:DOCPdPIC}, and the backoff terms are omitted, since the goal is to compute a nominal trajectory.

@@

\subsection{Step 3 -- Closed-loop robustified planning}
\label{sec:CLrobustified}
%\begin{align}
%    \underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k, \bK_k}{\text{minimize}} \quad & J_k(\bmu_k,\bxi_k, \bu_k) \label{eq:ROCPcost}\\
%       \text{s.t.}\notag\\
%          & f_0(x) - t \le 0 \quad \phantom{i=1, \dots, m}\\
%          & f_i(x) \le 0, \quad i=1, \dots, m \quad i=1, \dots, m\\
%          & h_i(x) = 0,   \quad i=1, \dots, p,
%\end{align}
As the third and last step, a closed-loop robustified planning problem is set up. Its aim is the definition of a mean trajectory of a stochastic system which, working in tandem with a state-feedback controller that aims at stabilizing it, is able to mitigate the perturbations and reduce the propagation of uncertainty. This helps to guarantee that the robustified path constraints can be satisfied over the entire planning horizon. The discretized version of this problem takes the form of the following NLP:
\begin{subequations}\label{eq:ROCP}
\begin{alignat}{3}
\underset{\bmu_k,\bxi_k, \bu_k, \bP_k,\bz_k, \bK_k}{\text{minimize}}
    & \quad J_k(\bmu_k,\bxi_k, \bu_k) & & \label{eq:ROCPcost} \\
\text{s.t.}\notag\\
\bzero      & = \; \bPsimu_k(\bmu_{k-1},\bmu_k,\bxi_k, \bu_k,\bz_k),
\quad k = 1,\ldots, N \label{eq:ROCPdyn} \\
\bmu_0      & = \; \bar{\bmu}_0 \label{eq:ROCPdynIC} \\
\bzero      & = \; \bPsiPCL_k(\bmu_k,\bxi_k, \bu_k, \tbP_{k-1},\tbP_k,\bSi_k,\bz_k, \bK_k),
\quad k = 1,\ldots, N \label{eq:ROCPdP} \\
\tbP_0       & = \; \bar{\bP}_0 \succeq 0 \label{eq:ROCPdPIC} \\
\bzero      & = \; \bOm_k(\bmu_k,\bxi_k, \bu_k,\bz_k), \quad k = 0,\ldots, N \label{eq:ROCPpath} \\
0    & \geq  h_i(\bmu_k, \bu_k, \bz_k) + \be_i(\bmu_k, \bu_k, \tbP_k, \bz_k), \quad k = 1,\ldots, N;\; \quad i \in \calI \label{eq:ROCPconstraints}\\
-\de \bK_k   & \leq \bK_k - \hbK_k \leq  \de \bK_k \label{eq:ROCPdeltaK}, \quad k = 0,\ldots, N
\end{alignat}
\end{subequations}
Here,~\eqref{eq:ROCPdP} denotes the discrete-time covariance dynamics under the stabilizing state-feedback controller \( \bbu_k = -\bK_k \bbmu_k \). To distinguish this case from~\eqref{eq:DOCP}, we use the symbol \( \tbP \) to denote the closed-loop (CL) covariance matrix, whose discrete evolution corresponds to a collocation-based approximation of the continuous-time dynamics $\dot{\tbP}(t) = \tbA(t) \tbP(t) + \tbP(t)\tbA(t)^T + \bQ(t)$, with $\tbA(t)=\bA(t)-\bB(t)\bK(t)$. It is worth noting that, accordingly, in equation~\eqref{eq:ROCPconstraints} the backoff term $\be_i$ is a function of closed-loop uncertainty $\tbP_k$. Another key aspect is that the state-feedback matrix $\bK_k$ is also treated as a decision variable. This reflects the fact that the driver may adjust his action to stabilize the new (to-be-planned) nominal trajectory using a policy that differs from the previous $\hbK_k$, computed in step (ii) via LQR. To promote convergence, however, $\bK_k$ is restrained within prescribed bounds ($\de \bK_k$) relative to $\hbK_k$. In our tests $\de \bK_k = 0.1 \hbK_k$.

@@

To correctly enforce the mutual exclusivity within the LQR synthesis via Riccati recursion, the state-feedback gain must be computed with respect to the minimal control input. Specifically, the control input matrix $\tbB$ is employed, and the gain $\tilde{\bK} \in \mathbb{R}^{2 \times 6}$, which operates on the independent variation \(\delta\tilde{\bu}\), is to be determined. This is structured as
\[
\tilde{\bK} = \begin{bmatrix} \tilde{\bK}_F^T & \tilde{\bK}_\delta^T \end{bmatrix}^T, \quad \tilde{\bK}_F, \tilde{\bK}_\delta \in \mathbb{R}^{1 \times 6}.
\]
The extended gain $\bK$ is automatically recovered via $\bK = \bH \tbK$ and, depending on the operating mode, it will result in the following expressions:
\begin{itemize}
  \item during \emph{acceleration}: \(\bK = \begin{bmatrix} \tilde{\bK}_F^T & \mathbf{0}^T & \tilde{\bK}_\delta^T \end{bmatrix}^T\),
  \item during \emph{braking}: \(\bK = \begin{bmatrix} \mathbf{0}^T & \tilde{\bK}_F^T & \tilde{\bK}_\delta^T \end{bmatrix}^T\).
\end{itemize}
This formulation preserves consistency with the complementarity constraint while retaining a structure amenable to controller synthesis.

### 2025-06-04 intro.tex
[mg] general reviews

In diverse fields, a variety of strategies has been proposed to tackle planning under uncertainty.                                          
In orbital mechanics, uncertainty in state estimation necessitates a probabilistic framework for modeling interactions and potential close approaches among natural and artificial celestial bodies.
A foundational reference in this context is~\cite{Tapley:StatisticalOrbitDetermination:2004}, where statistical orbit determination techniques are employed to assess and mitigate collision risks. In robotic motion planning, uncertainty-aware techniques are crucial to ensure safe navigation in environments populated with obstacles. One of the earliest contributions proposing a probabilistic representation of uncertainty is the chance-constrained framework in~\cite{Blackmore:ChanceConstrainedOptimalPath:2011}, which plans over the predicted distribution of the system state to ensure that the probability of constraint violation remains below a specified threshold. Within the Model Predictive Control paradigm, recent works have incorporated probabilistic safety guarantees. Notably, the methods presented in~\cite{Gao:CollisionfreeMotionPlanning:2023},~\cite{Zhang:RobustifiedTimeoptimalPointtopoint:2025}, and~\cite{Zhang:RobustifiedTimeoptimalCollisionfree:2024} introduce stochastic MPC frameworks for autonomous mobile platforms, where process noise is explicitly modeled and closed-loop tracking performance is maintained via either pre-computed or optimized feedback gains. The approach in~\cite{Gao:CollisionfreeMotionPlanning:2023} additionally proposes a zero-order optimization scheme to include the feedback gain directly in the optimal control problem, mitigating the growth of uncertainty over the planning horizon.
In the domain of chemical engineering, robust MPC approaches have been proposed to address uncertainty in industrial settings. For example,~\cite{Krog:SimpleFastRobust:2024} presents a heuristic method based on $n$-step-ahead uncertainty predictions, which are used to compute constraint tightening margins for the control of a polymerization reactor.

@@

Following an alternative approach, the largest Lyapunov exponent (LLE) has been proposed as an indicator of local stability~\cite{Meng:AnalysisGlobalCharacteristics:2022}, since it quantifies the average exponential rate at which infinitesimal perturbations off a fiducial trajectory grow or decay -- thus providing a direct measure of how deviations propagate through the system (negative values denote convergence toward the reference trajectory, while positive values indicate divergence)~\cite{Strogatz:NonlinearDynamicsChaos:2019}. 
%Except in chaotic systems where a positive LLE may still correspond to bounded motion on a strange attractor - a negative LLE typically signals local stability. 
Notably, for trajectories converging to a stable equilibrium, the (negative) LLE asymptotically approaches the real part of the largest eigenvalues of the linearized system at that point. This result has been leveraged in the literature to infer whether a system, not yet at equilibrium, is ultimately converging toward one, thereby enabling a localized assessment of stability based on the transient behavior~\cite{Sadri:StabilityAnalysisNonlinear:2013}. Applications of this method are particularly relevant to study chaotic vessel motions~\cite{McCue:UseLyapunovExponents:2011} and rotorcraft dynamics~\cite{Tamer:StabilityNonlinearTimeDependent:2016,Cassoni:RotorcraftStabilityAnalysis:2024}, where the primary concern is understanding the long-term evolution of the system and determining its asymptotic behavior.

%However, while Lyapunov-based indicators have been applied in vehicular contexts as well~\cite{Sadri:StabilityAnalysisNonlinear:2013,Meng:AnalysisGlobalCharacteristics:2022}, their relevance here is arguably of reduced interest - not due to their binary nature, but because even identifying asymptotic stability regions offers limited practical insight during planning. A trajectory may lie within the basin of attraction of a distant equilibrium, yet the long-term convergence may be of little operational use. What matters instead is whether, in the short term, a realistically modeled driver - acting in feedback or planning a bit into the future - can ensure that the vehicle's response to perturbations remains within safety-critical constraints, such as grip limits and collision avoidance. This perspective prioritizes short-horizon constraint-satisfying stabilizability over asymptotic notions of stability.
Although Lyapunov-based indicators have been applied in vehicular contexts as well~\cite{Sadri:StabilityAnalysisNonlinear:2013,Meng:AnalysisGlobalCharacteristics:2022}, their utility in planning is limited by the fact that asymptotic stability -- while theoretically informative -- often lacks operational relevance. A point along the planned trajectory may belong to the basin of attraction of a stable equilibrium, but if that equilibrium lies far ahead in time or outside the operational domain (e.g., off track), its practical relevance is questionable.

@@

\subsection{Paper's contributions and organization}
We tackle the challenge of embedding robustness against disturbances directly into minimum-lap-time planning and propose two complementary methods. In the first one -- the open-loop worst-case covariance propagation -- at each discretization point, we propagate the state covariance forward over a fixed horizon and tighten all path constraints against the maximum covariance growth. By back-offing constraints using this \emph{worst-case} covariance, the resulting trajectory maximizes robustness without relying on feedback during planning.

In the second one -- the closed-loop covariance-aware planning, we integrate a time-varying LQR feedback policy into the planning process to realistically tame uncertainty growth. First, a nominal time-optimal trajectory is computed under deterministic dynamics. Second, an LQR controller is designed to stabilize that nominal path. Finally, we re-solve the planning problem - together with the feedback gains - so that, along this new nominal trajectory, the closed-loop covariance propagation satisfies tightened constraints via a Lyapunov-based formulation. These two approaches offer a balance between computational simplicity and realistic, feedback-informed robustness, laying the groundwork for safe, high-performance lap-time planning under uncertainty in the motorsport context.

The rest of the paper is organized as follows. 
In Section~\ref{sec:svdf} we introduce our stochastic dynamics framework, recalling the single-track vehicle model with nonlinear tires used for planning, and formulate both the continuous-time planning problem and its discrete-time counterpart via direct-collocation. Here, we also derive the probabilistic constraint back-off formulation.
 Section~\ref{sec:open_loop_planning} then outlines the open-loop worst-case covariance-propagation planning approach, while Section~\ref{sec:closed_loop_planning} presents the closed-loop, covariance-aware planner. In Section~\ref{sec:performance_comparison} we compare both methods on a representative track scenario, reporting adhesion and collision-avoidance results, parameter-sensitivity studies, and performance trade-offs. Finally, Section~\ref{sec:conclusions} draws conclusions and offers directions for future work.

>>>>

In diverse fields, a variety of strategies has been proposed to tackle planning under uncertainty.                                          
In orbital mechanics, uncertainty in state estimation necessitates a probabilistic framework for modeling interactions and potential close approaches among natural and artificial celestial bodies.
A foundational reference in this context is~\cite{Tapley:StatisticalOrbitDetermination:2004}, where statistical orbit determination techniques are employed to assess and mitigate collision risks. In robotic motion planning, uncertainty-aware techniques are crucial to ensure safe navigation in environments populated with obstacles. One of the earliest contributions proposing a probabilistic representation of uncertainty is the chance-constrained framework in~\cite{Blackmore:ChanceConstrainedOptimalPath:2011}, which plans over the predicted distribution of the system state to ensure that the probability of constraint violation remains below a specified threshold.

Within the Model Predictive Control (MPC) paradigm, recent works have incorporated probabilistic safety guarantees. Notably, the methods presented in~\cite{Gao:CollisionfreeMotionPlanning:2023},~\cite{Zhang:RobustifiedTimeoptimalPointtopoint:2025}, and~\cite{Zhang:RobustifiedTimeoptimalCollisionfree:2024} introduce stochastic MPC frameworks for autonomous mobile platforms, where process noise is explicitly modeled and closed-loop tracking performance is maintained via either pre-computed or optimized feedback gains. The approach in~\cite{Gao:CollisionfreeMotionPlanning:2023} additionally proposes a zero-order optimization scheme to include the feedback gain directly in the optimal control problem, mitigating the growth of uncertainty over the planning horizon.
In the domain of chemical engineering, robust MPC approaches have been proposed to address uncertainty in industrial settings. For example,~\cite{Krog:SimpleFastRobust:2024} presents a heuristic method based on $n$-step-ahead uncertainty predictions, which are used to compute constraint tightening margins for the control of a polymerization reactor.

@@

Following an alternative approach, the largest Lyapunov exponent (LLE) has been proposed as an indicator of local stability~\cite{Meng:AnalysisGlobalCharacteristics:2022}. The LLE quantifies the average exponential rate at which infinitesimal perturbations off a fiducial trajectory grow or decay---thus providing a direct measure of how deviations propagate through the system. 
\textcolor{red}{(negative values denote convergence toward the reference trajectory, while positive values indicate divergence)~\cite{Strogatz:NonlinearDynamicsChaos:2019}. 
%Except in chaotic systems where a positive LLE may still correspond to bounded motion on a strange attractor - a negative LLE typically signals local stability. 
Notably, for trajectories converging to a stable equilibrium, the (negative) LLE asymptotically approaches the real part of the largest eigenvalues of the linearized system at that point. This result has been leveraged in the literature to infer whether a system, not yet at equilibrium, is ultimately converging toward one, thereby enabling a localized assessment of stability based on the transient behavior~\cite{Sadri:StabilityAnalysisNonlinear:2013}.} Applications of this method are particularly relevant to study chaotic vessel motions~\cite{McCue:UseLyapunovExponents:2011} and rotorcraft dynamics~\cite{Tamer:StabilityNonlinearTimeDependent:2016,Cassoni:RotorcraftStabilityAnalysis:2024}, where the primary concern is understanding the long-term evolution of the system and determining its asymptotic behavior.
%However, while Lyapunov-based indicators have been applied in vehicular contexts as well~\cite{Sadri:StabilityAnalysisNonlinear:2013,Meng:AnalysisGlobalCharacteristics:2022}, their relevance here is arguably of reduced interest - not due to their binary nature, but because even identifying asymptotic stability regions offers limited practical insight during planning. A trajectory may lie within the basin of attraction of a distant equilibrium, yet the long-term convergence may be of little operational use. What matters instead is whether, in the short term, a realistically modeled driver - acting in feedback or planning a bit into the future - can ensure that the vehicle's response to perturbations remains within safety-critical constraints, such as grip limits and collision avoidance. This perspective prioritizes short-horizon constraint-satisfying stabilizability over asymptotic notions of stability.
Although Lyapunov-based indicators have been applied in vehicular contexts as well~\cite{Sadri:StabilityAnalysisNonlinear:2013,Meng:AnalysisGlobalCharacteristics:2022}, their utility in planning is limited by the fact that asymptotic stability -- while theoretically informative -- often lacks operational relevance. A point along the planned trajectory may belong to the basin of attraction of a stable equilibrium, but if that equilibrium lies far ahead in time or outside the operational domain (e.g., off track), its practical relevance is questionable.

@@

\subsection{Paper's contributions and organization}
We tackle the challenge of embedding robustness against disturbances directly into minimum-lap-time planning and propose two complementary methods. In the first one---the open-loop \emph{horizon-based} covariance propagation---at each discretization point, we propagate the state covariance forward over a fixed horizon and tighten all path constraints against the maximum covariance growth. By back-offing constraints using this worst-case covariance, the resulting trajectory maximizes robustness without relying on feedback during planning.

In the second one -- the closed-loop covariance-aware planning, we integrate a time-varying LQR feedback policy into the planning process to realistically tame uncertainty growth. First, a nominal time-optimal trajectory is computed under deterministic dynamics. Second, an LQR controller is designed to stabilize that nominal path. Finally, we re-solve the planning problem - together with the feedback gains - so that, along this new robust trajectory, the closed-loop covariance propagation satisfies tightened constraints via a Lyapunov-based formulation. These two approaches offer a balance between computational simplicity and realistic, feedback-informed robustness, laying the groundwork for safe, high-performance lap-time planning under uncertainty in the motorsport context.

The rest of the paper is organized as follows. 
In Section~\ref{sec:svdf} we introduce our stochastic dynamics framework, recalling the single-track vehicle model with nonlinear tires used for planning, and formulate both the continuous-time planning problem and its discrete-time counterpart via direct-collocation. Here, we also derive the probabilistic constraint back-off formulation.
Section~\ref{sec:open_loop_planning} then outlines the open-loop horizon-based covariance-propagation planning approach, while Section~\ref{sec:closed_loop_planning} presents the closed-loop, covariance-aware planner. In Section~\ref{sec:performance_comparison} we compare both methods on a representative track scenario, reporting parameter-sensitivity studies, analyzing the influence of different sets of constraints, and performance trade-offs. 
Finally, Section~\ref{sec:conclusions} draws conclusions and offers directions for future work.

### 2025-06-04 open_loop_planning.tex
[mg] general reviews

\section{Open-loop planning via H-steps ahead predictions}
\label{sec:open_loop_planning}
In this section, we address the problem of planning under uncertainty for a stochastic system, where the covariance matrix is propagated in a feed-forward fashion for a fixed prediction horizon of $H$ steps at each grid point $k$ of the discretized trajectory. In the planning problem, we conservatively evaluate the maximum propagation of the covariance, that is, the matrix evolved over $H$ steps, to ensure the tightest possible enforcement of constraints. This maximizes the robustness of the resulting solution, as the constraint backoff is evaluated in the worst-case scenario.

@@

The repeated initialization of the covariance matrix at each step is necessary to prevent its unbounded growth, as we assume that the system evolves without feedback control.
In the absence of feedback, the uncertainty on the position and orientation of the body frame accumulates along the track, which may severely and unrealistically degrade the conditioning of the covariance matrix. 
For this reason, the propagation horizon in this open-loop setting is a critical design parameter that must be bounded to realistically capture the evolution of disturbances occurring prior to any corrective driver response.

>>>>

\section{Open-loop planning via H-steps ahead predictions}
\label{sec:open_loop_planning}
In this section, we address the problem of planning under uncertainty for a stochastic system. 
At each grid point $k$ of the discretized trajectory the covariance matrix is propagated in a feed-forward fashion for a fixed prediction horizon of $H$ steps.
In the planning problem, we conservatively evaluate the maximum propagation of the covariance, that is, the matrix evolved over $H$ steps, to ensure the tightest possible enforcement of constraints. This maximizes the robustness of the resulting solution, as the constraint back-off is evaluated in the worst-case scenario within the prediction horizon.

@@

The repeated initialization of the covariance matrix at each step is necessary to prevent its unbounded growth, as we assume that the system evolves without feedback control.
In the absence of feedback, the uncertainty on the position and orientation of the body frame accumulates along the track, leading to a severe and unrealistic degradation of the covariance matrix conditioning.
For this reason, the propagation horizon in this open-loop setting is a critical design parameter.
We bound the number of steps $H$ to realistically capture the evolution of disturbances occurring prior to any corrective driver response.
